---
title: "Fertility Modelling and Sibling Estimates"
author: "Junhui He"
date: today
---

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE, fig.width = 9, fig.height = 7)

library(dplyr)
library(tidyr)
library(readr)
library(MASS)     ## glm.nb
library(pscl)     ## zeroinfl
library(knitr)    ## kable
library(ggplot2)
library(viridisLite)
```

# IPMUS fertility data

We extract the fertility information for U.S. women from the Integrated Public Use Microdata Series (IPUMS) dataset.

## Load the IPUMS subset

```{r load-data}
path         <- if (dir.exists("data")) "data" else file.path("..", "data")
output_path  <- if (dir.exists("output")) "output" else file.path("..", "output")
file_path    <- file.path(path, "mother_data.csv")

## original script loads both; we only need mother_data for this step
mother_data <- read_csv(file_path, show_col_types = FALSE)
```

`mother_data` already contains:

-   `Year` – Census year (1960,1970,1980,1990)
-   `Age` – respondent age
-   `Race` – self-identified race of the respondent.
-   `n_child` – the number of children ever born to each woman

## Prepare the fertility data by generations

We focus on U.S. women aged 50-59, who are considered to have completed their fertility. The Census data include observations from 1960, 1970, 1980, and 1990. Women who were aged 50-59 in 1950 are observed as ages 60-69 in the 1960 Census. We therefore prepare fertility data for U.S. women aged 50-59 in 1950, 1960, 1970, 1980, and 1990, corresponding to the 1891-1900, 1901-1910, 1911-1920, 1921-1930, and 1931-1940 birth cohorts, respectively.

```{r prepare-fertility-data}
fertility_data <- bind_rows(
  mother_data %>%
    filter(Year %in% c(1960, 1970, 1980, 1990),
           Age >= 50, Age <= 59),
  mother_data %>%
    filter(Year == 1960,
           Age >= 60, Age <= 69) %>%
    transmute(
      Year = 1950,
      Age = Age - 10,
      Race = Race,
      n_child = n_child
    )
) %>%
  filter(!is.na(n_child)) %>%
  arrange(Year, Age)

fertility_data %>%
  count(Year, name = "n_women") %>%
  kable(caption = "Prepared fertility data by census year and age.")

head(fertility_data, n = 15) %>%
  kable(caption = "Fertility data for U.S. women aged 50‑59 in the 1950, 1960, 1970, 1980, and 1990 Censuses.")
```

## Prepare individual-level sibling counts

```{r prepare-sibling-counts}
# Prepare individual-level sibling count
sib_data <- fertility_data %>%
  mutate(mother_id = row_number()) %>%
  filter(n_child > 0) %>%
  tidyr::uncount(n_child, .remove = FALSE) %>%
  mutate(n_sib = n_child - 1)

kable(head(sib_data, n = 15), caption = "Corresponding sibling data calculated from fertility data")
```

## Visualise the fertility and sibling distributions across Census years

```{r visualize-fertility-sibling, fig.width = 11, fig.height = 7}
plot_census_pmf <- function(fertility_df, sibling_df, max_count = 12) {
  pmf_df <- bind_rows(
    fertility_df %>%
      transmute(
        Year,
        Distribution = "Fertility",
        count = pmin(n_child, max_count)
      ),
    sibling_df %>%
      transmute(
        Year,
        Distribution = "Sibling",
        count = pmin(n_sib, max_count)
      )
  ) %>%
    count(Year, Distribution, count, name = "n_counts") %>%
    group_by(Year, Distribution) %>%
    mutate(prob = n_counts / sum(n_counts)) %>%
    ungroup()
  
  count_levels <- 0:max_count
  count_labels <- as.character(count_levels)
  count_labels[length(count_labels)] <- paste0(max_count, "+")
  
  pmf_df %>%
    mutate(
      signed_prob = if_else(Distribution == "Fertility", -prob, prob),
      count_cat = factor(count, levels = count_levels, labels = count_labels),
      Year = factor(Year)
    ) %>%
    ggplot(aes(x = count_cat, y = signed_prob, fill = count_cat, alpha = Distribution)) +
    geom_col(width = 0.82) +
    geom_hline(yintercept = 0, color = "black", linewidth = 0.4) +
    coord_flip() +
    facet_wrap(~ Year, ncol = 3) +
    scale_y_continuous(labels = function(x) abs(x)) +
    scale_fill_manual(values = viridis(length(count_levels), option = "D"), name = "Count") +
    scale_alpha_manual(values = c("Fertility" = 0.7, "Sibling" = 1), guide = "none") +
    labs(
      #title = "Probability mass functions (PMF) of fertility and sibling distributions across years",
      x = "Count",
      y = "Probability",
      # caption = "Left: Fertility distribution; Right: Sibling distribution"
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(size = 14, hjust = 0.5),
      axis.text.y = element_text(size = 10),
      plot.caption = element_text(face = "bold", size = 10),
      legend.position = "none",
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank()
    )
}

plot_census_pmf(fertility_data, sib_data)
ggsave(file = file.path(output_path, "census_fertility_sibling_pmf.pdf"), width = 11, height = 7)
```

```{r plot-mean-across-years, fig.width = 8, fig.height = 5}
mean_across_years <- bind_rows(
  fertility_data %>%
    group_by(Year) %>%
    summarise(
      Distribution = "Fertility",
      Mean = mean(n_child),
      .groups = "drop"
    ),
  sib_data %>%
    group_by(Year) %>%
    summarise(
      Distribution = "Sibling",
      Mean = mean(n_sib),
      .groups = "drop"
    )
)

mean_across_years %>%
  ggplot(aes(x = Year, y = Mean, color = Distribution, group = Distribution)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = sort(unique(mean_across_years$Year))) +
  scale_color_manual(
    values = c("Fertility" = "#0072B2", "Sibling" = "#D55E00"),
    name = NULL
  ) +
  labs(
    # title = "Mean Fertility and Sibling Counts Across Census Years",
    x = "Year",
    y = "Mean family size"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom"
  )
ggsave(file = file.path(output_path, "mean_fertility_sibling_across_years.pdf"), width = 8, height = 5)
```


# Family structure modelling

Here we establish statistical models for the **fertility distribution** and **sibling distribution**. The fertility distribution describes the number of children per family, while the sibling distribution describes the number of siblings per person.

## Fertility distribution

We model the number of children per family using a **Poisson distribution** and a **zero inflated negative binomial distribution (ZINB)**, respectively.

### Poisson distribution

The **Poisson distribution** is a common model for count data, particularly when the mean and variance are equal. However, it may not be suitable for fertility data, which often exhibit zero-inflation and overdispersion (variance greater than the mean). The Poisson distribution with **mean fertility** ($\mu$) is defined as: for a count $k \geq 0$, $$P(num_{child}=k)=e^{-\mu}\frac{\mu^k}{k!},$$ denoted by $Poisson(\mu)$. The mean and variance of $Poisson(\mu)$ are both $\mu$.

### Zero inflated negative binomial distribution

The **zero inflated negative binomial distribution** with parameters **mean fertility** ($\mu$), **dispersion/size** ($\theta$), and **zero‑inflation probability** ($\pi_0$)---denoted as $ZINB(\mu, \theta, \pi_0)$---is defined as: for a count $k \geq 0$, $$P(num_{child}=k)=\begin{cases}
\pi_0 + (1-\pi_0)p^{\theta} & \text{if } k=0, \\
(1-\pi_0)\binom{k+\theta-1}{k}(1-p)^k p ^{\theta} & \text{if } k \geq 1,
\end{cases}$$ where $p = \frac{\theta}{\mu+\theta}$ represents the success probability of the negative binomial distribution.

According to the definition of the negative binomial distribution, the mean and variance of $NB(\mu,\theta)$ are given by $\mu$ and $\mu + \frac{\mu^2}{\theta}$, respectively. Therefore, the mean and variance of $ZINB(\mu, \theta, \pi_0)$ are defined by $\mu(1-\pi_0)$ and $\mu(1-\pi_0) + \frac{\mu^2(1-\pi_0)}{\theta}+\mu^2\pi_0(1-\pi_0)$, respectively.

## Sibling distribution

The **sibling distribution** can be derived from the fertility distribution. For any count $k \geq 0$, the probability that an individual has $k$ siblings is proportional to the probability that their mother has $k+1$ children, multiplied by the number of ways the individual could be one of those children. Therefore, the sibling distribution satisfies: $$P(num_{sib}=k) \propto (k+1)P(num_{child}=k+1).$$

### Derivation of sibling distribution

If the **fertility distribution** is $Poisson(\mu)$, then the **sibling distribution** is given by: for a count $k \geq 0$, $$P(num_{sib}=k) \propto (k+1)e^{-\mu}\frac{\mu^{k+1}}{(k+1)!} = e^{-\mu}\frac{\mu^{k+1}}{k!}\propto e^{-\mu}\frac{\mu^k}{k!},$$ which is also a **Poisson distribution** with mean $\mu$.

If the **fertility distribution** is $ZINB(\mu, \theta, \pi_0)$, then the **sibling distribution** is given by: for a count $k \geq 0$, $$\begin{align*}
P(num_{sib}=k) & \propto (1-\pi_0)(k+1)\binom{k+\theta}{k+1}(1-p)^{k+1} p^{\theta} \\
& = (1-\pi_0)(k+1)\frac{(k+\theta)!}{(k+1)!(\theta-1)!}(1-p)^{k+1} p^{\theta} \\
& \propto \frac{\theta(1-p)}{p} \frac{(k+\theta)!}{k!\theta!}(1-p)^k p^{\theta+1} \\
& \propto \binom{k+\theta}{k}(1-p)^k p^{\theta+1}, \\
\end{align*}$$ where $p = \frac{\theta}{\mu+\theta}$. Therefore, the sibling distribution is a **negative binomial distribution** with parameters **mean fertility** ($\frac{\mu(\theta+1)}{\theta}$), **dispersion/size** ($\theta+1$), denoted as $NB(\frac{\mu(\theta+1)}{\theta}, \theta+1)$. The variance is $\frac{\mu(\theta+1)}{\theta} + \frac{\mu^2(\theta+1)}{\theta^2}$.


# Model fitting

## Fit the Poisson and ZINB models

```{r fit-models}
fit_fert <- function(dat) {
  ## Poisson model
  pois <- glm(n_child ~ 1, family = poisson, data = dat)
  
  ## Zero‑inflated negative binomial model
  zinb <- zeroinfl(n_child ~ 1 | 1, data = dat, dist = "negbin")
  
  tibble(
    Model = c("Poisson", "ZINB"),
    mu    = c(exp(coef(pois)[1]),                    ## log‑link → exp() gives mean
              exp(zinb$coefficients$count[1])),
    size  = c(Inf,                                   ## dispersion θ; Inf = Poisson
              zinb$theta),
    pi0   = c(0,                                      ## zero‑inflation prob.
              plogis(zinb$coefficients$zero[1]))
  )
}

param_fert <- fertility_data %>%
  group_by(Year) %>%
  group_modify(~ fit_fert(.x)) %>%
  ungroup()
```

## Estimated fertility parameters

```{r fertility-params}
kable(param_fert, digits = 3, caption = "Fertility‑model parameter estimates for women aged 50‑59 (1950, 1960, 1970, 1980, 1990)")
```

## Compare the mean and variance from the models and data

```{r compare-mean-var}
cal_mean_var <- function(model, mu, size, pi0) {
  if (model == "Poisson") {
    mean_fert = mu
    var_fert  = mu
    mean_sib = mu
    var_sib  = mu
  } else if (model == "ZINB") {
    mean_fert = mu * (1 - pi0)
    var_fert  = mu * (1 - pi0) + (mu^2 * (1 - pi0) / size) + (mu^2 * pi0 * (1 - pi0))
    mean_sib = mu * ((size + 1) / size)
    var_sib  = mu * ((size + 1) / size) + (mu^2 * ((size + 1) / size) / size)
  }
  
  tibble(
    Model = model,
    Mean_Fertility = mean_fert,
    Variance_Fertility = var_fert,
    Mean_Sibling = mean_sib,
    Variance_Sibling = var_sib
  )
}

mean_var_fitted <- param_fert %>%
  rowwise(Year, Model) %>%
  reframe(cal_mean_var(Model, mu, size, pi0)) %>%
  ungroup()

mean_var_data <- bind_rows(
  fertility_data %>%
    group_by(Year) %>%
    summarise(
      distribution = "fertility",
      mean_data = mean(n_child),
      variance_data = var(n_child),
      .groups = "drop"
    ),
  sib_data %>%
    group_by(Year) %>%
    summarise(
      distribution = "sibling",
      mean_data = mean(n_sib),
      variance_data = var(n_sib),
      .groups = "drop"
    )
)

mean_var_model <- mean_var_fitted %>%
  transmute(
    Year,
    Model = tolower(Model),
    mean_fertility = Mean_Fertility,
    variance_fertility = Variance_Fertility,
    mean_sibling = Mean_Sibling,
    variance_sibling = Variance_Sibling
  ) %>%
  pivot_longer(
    cols = c(mean_fertility, variance_fertility, mean_sibling, variance_sibling),
    names_to = c(".value", "distribution"),
    names_pattern = "(mean|variance)_(fertility|sibling)"
  ) %>%
  pivot_wider(
    names_from = Model,
    values_from = c(mean, variance),
    names_glue = "{.value}_{Model}"
  )

mean_var <- mean_var_data %>%
  left_join(mean_var_model, by = c("distribution", "Year")) %>%
  dplyr::select(
    distribution,
    Year,
    mean_data,
    mean_poisson,
    mean_zinb,
    variance_data,
    variance_poisson,
    variance_zinb
  ) %>%
  arrange(distribution, Year)

write.csv(mean_var, file.path(output_path, "fertility_estimation.csv"), row.names = FALSE)

kable(mean_var, digits = 3, caption = "Mean and variance of fertility and sibling distributions from the fitted models and data.")
```

## Model evaluation

Based on the table of mean and variance, we can see that the fertility data exhibits **overdispersion (variance greater than mean)**. The **Poisson Model** assumes equal mean and variance, which does not fit the data well. The **ZINB Model** accounts for overdispersion and zero-inflation, providing a better fit to the data. The ZINB model's mean and variance estimates for both fertility and sibling distributions align more closely with the observed data.

# Simulations of fertility and sibling distributions

In this section, we simulate the fertility and sibling counts based on the Poisson and ZINB distributions, respectively. We then compare the empirical mean and variance of the simulated sibling data with the theoretical values derived from Section 2.

## Simulate the fertility and sibling counts

```{r simulate-fert-sib}
simulate_fert_sib <- function(model, mu, size, pi0, n_sim = 20000, max_kids = 12) {
  if (model == "Poisson") {
    # Simulate fertility counts from Poisson distribution
    fert_counts <- pmin(rpois(n_sim, lambda = mu), max_kids)
  } else if (model == "ZINB") {
    # Simulate fertility counts from ZINB distribution
    fert_counts <- ifelse(runif(n_sim) < pi0, 0, pmin(rnbinom(n_sim, size, mu = mu), max_kids))
  }
  
  # Simulate sibling counts based on fertility counts
  fert_counts_tibble <- tibble(fert_counts = fert_counts)
  sib_counts_tibble <- fert_counts_tibble %>%
    filter(fert_counts > 0) %>%
    tidyr::uncount(fert_counts, .remove = FALSE) %>%
    mutate(sib_counts = fert_counts - 1)
  sib_counts <- sib_counts_tibble$sib_counts
  
  list(
    model = model,
    fert_counts = fert_counts,
    sib_counts = sib_counts
  )
}
```

## Compare sibling counts from simulations with theoretical values

We run simulations for fertility means of 1.5, 2.0, 2.5, 3.0, 3.5, and 4.0 for both Poisson and ZINB models. Additionally, in ZINB models, we set the dispersion parameter $\theta$ to 4, and the zero-inflation probability $\pi_0$ to 0.05. 

We compare the empirical mean and variance of sibling counts from the simulated data with the theoretical values. The comparison shows that **the theoretical values align closely with the simulated results**, confirming the correctness of our derivation.

```{r compare-simulated-sibling}
means = c(1.5, 2.0, 2.5, 3.0, 3.5, 4.0)
size = 4
pi0 = 0.05

cal_mean_var_simulated <- function(sim_data) {
  sib_counts <- sim_data$sib_counts
  
  mean_sib = mean(sib_counts)
  var_sib = var(sib_counts)
  
  tibble(
    Mean_Sibling = mean_sib,
    Variance_Sibling = var_sib
  )
}

# Poisson Model
simulated_poisson <- lapply(means, function(mu) {
  simulate_fert_sib("Poisson", mu, Inf, 0)
})

simulated_poisson_results <- lapply(simulated_poisson, cal_mean_var_simulated) %>% bind_rows() %>%
  mutate(Fertility_Parameter = sapply(means, function(mu) paste0("Fertility Mean = ", mu)),
         Value_Type = "Simulated Poisson") %>% dplyr::select(Fertility_Parameter, Value_Type, everything())
threo_possion_results <- tibble(
  Fertility_Parameter = sapply(means, function(mu) paste0("Fertility Mean = ", mu)),
  Value_Type = "Theoretical Poisson",
  Mean_Sibling = means,
  Variance_Sibling = means
)

poisson_results <- bind_rows(simulated_poisson_results, threo_possion_results) %>%
  arrange(Fertility_Parameter) %>%
  mutate(Value_Type = factor(Value_Type, levels = c("Theoretical Poisson", "Simulated Poisson")))

kable(poisson_results, digits = 3, caption = "Mean and Variance of Sibling counts from simulations and theoretical values based on Poisson distribution.")

# ZINB Model
simulated_zinb <- lapply(means, function(mu) {
  simulate_fert_sib("ZINB", mu/(1-pi0), size, pi0)
})
simulated_zinb_results <- lapply(simulated_zinb, cal_mean_var_simulated) %>% bind_rows() %>%
  mutate(Fertility_Parameter = sapply(means, function(mu) paste0("Fertility Mean = ", mu)),
         Value_Type = "Simulated ZINB") %>% dplyr::select(Fertility_Parameter, Value_Type, everything())
mean_var_zinb <- lapply(means, function(mu) cal_mean_var("ZINB", mu/(1-pi0), size, pi0)) %>% bind_rows()
threo_zinb_results <- tibble(
  Fertility_Parameter = sapply(means, function(mu) paste0("Fertility Mean = ", mu)),
  Value_Type = "Theoretical ZINB",
  Mean_Sibling = mean_var_zinb$Mean_Sibling,
  Variance_Sibling = mean_var_zinb$Variance_Sibling
)

zinb_results <- bind_rows(simulated_zinb_results, threo_zinb_results) %>%
  arrange(Fertility_Parameter) %>%
  mutate(Value_Type = factor(Value_Type, levels = c("Theoretical ZINB", "Simulated ZINB")))

kable(zinb_results, digits = 3, caption = "Mean and Variance of Sibling counts from simulations and theoretical values based on ZINB distribution.")
```

## Visualize sibling distributions

We generate discrete probability distributions for sibling counts based on both Poisson and ZINB models, with the y-axis representing the number of siblings and the x-axis representing the probability of having that many siblings. The fertility mean is set to 2.5, and the ZINB model has a dispersion parameter of 4 and a zero-inflation probability of 0.05.

```{r plot-pmf}
library(viridisLite)

plot_pmf <- function(pmf_df) {
  cats = sort(unique(pmf_df$sib_counts))
  n_cat <- length(cats)
  color_palette <- viridis(n_cat, option = "D")
  df <- pmf_df %>%
    mutate(y = ifelse(Model == "Poisson", -prob, prob),
           count_cat = factor(sib_counts, levels = cats))
  ggplot(df, aes(x = count_cat, y = y, fill = count_cat, alpha = Model)) +
    geom_col() +
    geom_hline(yintercept = 0, color = "black", size = 0.5) +
    coord_flip() +
    scale_y_continuous(labels = function(x) abs(x)) +
    scale_fill_manual(values = color_palette, name = "Siblings Count") +
    scale_alpha_manual(values = c("Poisson" = 0.7, "ZINB" = 1), guide = "none") +
    labs(
      title = "PMF of Siblings Counts (Simulated from Poisson and ZINB Models)",
      x = "Number of Siblings",
      y = "Probability",
      caption = "Poisson (left, negative y), ZINB (right, positive y)."
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(size = 14, hjust = 0.5),
      axis.text.y = element_text(size = 10),
      plot.caption = element_text(face = "bold", size = 10),
      legend.position = "right",
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank()
    )
}
```


```{r visualize-sibling-distributions}
# fertility parameters
mean = 2.5
size = 4
pi0 = 0.05
max_sibs = 12

# simulated data for fertility mean 2.5
simulated_data <- bind_rows(
  tibble(sib_counts = simulate_fert_sib("Poisson", mean, Inf, 0)$sib_counts) %>%
    mutate(Model = "Poisson"),
  tibble(sib_counts = simulate_fert_sib("ZINB", mean/(1-pi0), size, pi0)$sib_counts) %>%
    mutate(Model = "ZINB")
)

simulated_data$sib_counts <- ifelse(simulated_data$sib_counts > max_sibs, max_sibs, simulated_data$sib_counts)

pmf_data <- simulated_data %>%
  count(Model, sib_counts, name = "n_counts") %>%
  group_by(Model) %>%
  mutate(prob = n_counts / sum(n_counts)) %>%
  ungroup()

# write the pmf data to a csv file
write.csv(pmf_data, file = file.path(output_path, "pmf_sibling_counts.csv"), row.names = FALSE)
```

```{r plot-pmf-fig, fig.width = 8, fig.height = 6}
plot_pmf(pmf_data)
ggsave(file = file.path(output_path, "pmf_sibling_counts.png"), width = 8, height = 6, dpi = 500)
```

```{r}
sum(pmf_data$n_counts[pmf_data$Model == "Poisson"])
sum(pmf_data$n_counts[pmf_data$Model == "ZINB"])

pmf_data$n_counts[pmf_data$Model == "Poisson"] / sum(pmf_data$n_counts[pmf_data$Model == "Poisson"])
pmf_data$n_counts[pmf_data$Model == "ZINB"] / sum(pmf_data$n_counts[pmf_data$Model == "ZINB"])
```
