---
title: "Preprocessing IPUMS fertility data"
author: "Tina Lasisi"
date: today
execute:
  echo: true
  warning: false
  message: false
---

This notebook converts a restricted IPUMS USA extract into the aggregate fertility
summaries used throughout the kincounts project, and documents the processing
decisions. It makes two analytic contributions:

1. **Completed-fertility distributions by birth cohort** — women aged 50–59 in each
   census from 1950 to 1990 (children ever born), the inputs used by the downstream
   fertility model and simulator.
2. **A common-universe analysis** — a measurement of how those distributions change
   when the population that was asked the children-ever-born question shifts from
   *ever-married women* (1950–1960) to *all women* (1970–1990).

Because IPUMS USA policy restricts redistribution of raw microdata, only the aggregate
CSVs in `data/processed/` are committed; the raw extract stays local-only (ignored by
Git via `.gitignore`). Processing code is collapsed by default; the sections below
present the resulting figures and tables.

## Data source and access

The source data are an IPUMS USA extract (a fixed-width `*.dat.gz` plus its `*.xml` DDI
codebook) covering the 1950, 1960, 1970, 1980, and 1990 U.S. Census samples. The
variables used are `YEAR`, `PERWT`, `SEX`, `AGE`, `BIRTHYR`, `MARST`, `GQ`, and
`CHBORN`. The sample-line weight `SLWT`, the allocation flag `QCHBORN`, and the
nativity/birthplace variables `NATIVITY`/`BPL` were not requested; their absence is
discussed in the limitations.

Formal citation:

Steven Ruggles, Sarah Flood, Matthew Sobek, Daniel Backman, Grace Cooper, Julia A.
Rivera Drew, Stephanie Richards, Renae Rodgers, Jonathan Schroeder, and Kari C.W.
Williams. IPUMS USA: Version 16.0 [dataset]. Minneapolis, MN: IPUMS, 2025.
https://doi.org/10.18128/D010.V16.0

To reproduce the raw-data step, create your own IPUMS USA extract with the same samples
and variables, place the `.dat.gz` and `.xml` files in `data/raw/`, and render this
notebook locally. Only the aggregate CSVs in `data/processed/` should be committed.

## Data processing

By default, the code below loads the committed aggregate CSVs in `data/processed/`
without looking for or opening the restricted raw IPUMS files. To rebuild those
aggregates from a local raw extract, set `KINCOUNTS_REBUILD_IPUMS=true` for the render;
only that explicit flag triggers raw parsing.

```{r setup-data-access}
# Packages for IPUMS parsing, data manipulation, CSV I/O, tables, and figures.
library(ipumsr)
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(ggplot2)
library(viridisLite)

theme_set(theme_minimal(base_size = 12))

coverage_palette <- setNames(
  viridis(3, option = "D", begin = 0.12, end = 0.88),
  c("Ever-married, 1-in-6 sample line", "Ever-married women", "All women")
)
universe_palette <- setNames(
  viridis(2, option = "D", begin = 0.2, end = 0.85),
  c("All women", "Ever-married")
)
validation_palette <- setNames(
  viridis(2, option = "D", begin = 0.2, end = 0.85),
  c("Direct 1950", "Proxy (1960)")
)
main_series_color <- viridis(1, option = "D", begin = 0.55, end = 0.55)

# Resolve paths whether rendered from the project root or the analysis directory.
project_root <- if (dir.exists("data")) "." else ".."
raw_dir <- Sys.getenv("KINCOUNTS_RAW_DIR", file.path(project_root, "data", "raw"))
processed_dir <- file.path(project_root, "data", "processed")
dir.create(processed_dir, recursive = TRUE, showWarnings = FALSE)

processing_version <- "common_universe_v5"

# Variables parsed from the restricted person-level data when rebuilding.
analysis_vars <- c("YEAR", "PERWT", "SEX", "AGE", "BIRTHYR", "MARST", "GQ", "CHBORN")
ddi_display_vars <- analysis_vars

# Every aggregate file this notebook writes or expects to load.
output_inventory <- bind_rows(
  tibble(
    file = c(
      "ipums_preprocess_metadata.csv",
      "ipums_cohort_sample_sizes.csv",
      "ipums_fertility_counts_by_cohort.csv",
      "ipums_fertility_pmf_by_cohort.csv",
      "ipums_fertility_counts_by_cohort_weighted.csv",
      "ipums_fertility_pmf_by_cohort_weighted.csv",
      "validation_1950_direct_vs_proxy.csv",
      "validation_1950_direct_vs_proxy_pmf.csv",
      "ipums_age_cohort_validation.csv",
      "ipums_chborn_universe_by_cohort.csv",
      "ipums_cohort_fertility_checks.csv",
      "ipums_cohort_universe_summary.csv",
      "ipums_cohort_universe_pmf.csv",
      "ipums_universe_contrast_by_cohort.csv",
      "ipums_early_cohort_followup.csv"
    ),
    contents = c(
      "Processing version and rebuild timestamp.",
      "Main-series sample sizes by output cohort.",
      "Unweighted main-series fertility counts by cohort and child-count bucket.",
      "Unweighted main-series fertility PMF by cohort and child-count bucket.",
      "PERWT-weighted main-series fertility counts.",
      "PERWT-weighted main-series fertility PMF.",
      "Aggregate comparison of proxy 1950 and direct 1950 validation records.",
      "Child-count PMFs for proxy 1950 and direct 1950 validation records.",
      "Age range and derived birth-year span for each age-defined cohort source.",
      "Valid versus not-in-universe CHBORN counts by cohort source.",
      "Aggregate mean children and 12+ share by cohort source.",
      "Common-universe analysis: per decade-cohort x census x universe summary.",
      "Common-universe analysis: child-count PMFs per decade-cohort x census x universe.",
      "All-women vs ever-married contrast (mean shift, share-0 shift, TVD) by cohort.",
      "Early-cohort decomposition of universe vs mortality/decade effects."
    ),
    role = c(
      "Reproducibility check",
      "Downstream model input",
      "Downstream model input",
      "Downstream model input",
      "Sensitivity output",
      "Sensitivity output",
      "Validation output",
      "Validation output",
      "QA output",
      "QA output",
      "QA output",
      "Common-universe analysis",
      "Common-universe analysis",
      "Common-universe analysis",
      "Common-universe analysis"
    )
  ),
  tibble(
    file = paste0("fertility_pmf_", c(1950, 1960, 1970, 1980, 1990), ".csv"),
    contents = "One unweighted main-series fertility PMF for the named output year.",
    role = "Downstream simulator input"
  )
)

processed_files <- output_inventory$file
processed_paths <- file.path(processed_dir, processed_files)

force_rebuild <- tolower(Sys.getenv("KINCOUNTS_REBUILD_IPUMS", "false")) %in%
  c("true", "1", "yes")

dat_gz_file <- NA_character_
ddi_file <- NA_character_
has_raw_extract <- FALSE
if (force_rebuild) {
  dat_gz_files <- Sys.glob(file.path(raw_dir, "*.dat.gz"))
  ddi_files <- Sys.glob(file.path(raw_dir, "*.xml"))
  has_raw_extract <- length(dat_gz_files) == 1 && length(ddi_files) == 1
  if (!has_raw_extract) {
    stop(
      "KINCOUNTS_REBUILD_IPUMS=true requires exactly one .dat.gz and one .xml file in ",
      raw_dir,
      call. = FALSE
    )
  }
  dat_gz_file <- dat_gz_files
  ddi_file <- ddi_files
}

# Normal renders use the committed aggregate files. Rebuilding from the restricted
# person-level extract happens only when explicitly requested.
metadata_path <- file.path(processed_dir, "ipums_preprocess_metadata.csv")
metadata_current <- file.exists(metadata_path) &&
  identical(
    read_csv(metadata_path, show_col_types = FALSE)$processing_version[[1]],
    processing_version
  )
processed_complete <- all(file.exists(processed_paths)) && metadata_current
processed_current <- processed_complete

if (!force_rebuild && !processed_complete) {
  stop(
    "The processed CSV cache is incomplete or was built with an older processing version. ",
    "Use the committed data/processed files, or add the raw extract and render with ",
    "KINCOUNTS_REBUILD_IPUMS=true to rebuild.",
    call. = FALSE
  )
}

rebuild_processed <- force_rebuild

if (rebuild_processed) {
  # The DDI supplies the fixed-width layout and the value labels used below.
  ddi <- read_ipums_ddi(ddi_file)

  ddi_vars <- ipums_var_info(ddi)
  missing_vars <- setdiff(ddi_display_vars, ddi_vars$var_name)
  if (length(missing_vars) > 0) {
    stop(
      "The DDI is missing required variables: ",
      paste(missing_vars, collapse = ", "),
      call. = FALSE
    )
  }

  # Check that ipumsr can parse a few records using the DDI layout.
  parse_check <- read_ipums_micro(
    ddi,
    vars = analysis_vars,
    n_max = 5,
    data_file = dat_gz_file,
    verbose = FALSE
  )
  stopifnot(nrow(parse_check) == 5)
  rm(parse_check)

  # DDI variable positions (rendered in the appendix).
  ddi_variables <- ddi_vars %>%
    filter(var_name %in% ddi_display_vars) %>%
    arrange(match(var_name, ddi_display_vars)) %>%
    select(var_name, var_label, start, end, imp_decim)

  # DDI value labels that drive the coding rules (rendered in the coding section).
  value_labels <- bind_rows(lapply(c("SEX", "CHBORN", "MARST"), function(var) {
    labels <- ddi_vars$val_labels[[match(var, ddi_vars$var_name)]]
    mutate(labels, var_name = var, .before = 1)
  }))
} else {
  ddi_variables <- tibble(
    var_name = ddi_display_vars,
    var_label = "Raw DDI unavailable in processed-only render",
    start = NA_integer_,
    end = NA_integer_,
    imp_decim = NA_integer_
  )
  value_labels <- tibble(
    var_name = c("SEX", "SEX", "CHBORN", "CHBORN", "CHBORN", "CHBORN", "MARST", "MARST", "MARST"),
    val = c("1", "2", "0", "1", "13", "97-99", "1-5", "6", "9"),
    lbl = c(
      "Male",
      "Female",
      "N/A (not in universe)",
      "No children",
      "12 children; 12+ in 1960-1990",
      "Unknown, illegible, or missing",
      "Ever-married",
      "Never-married",
      "Blank/missing"
    )
  )
}

# Local filenames and cache state (rendered in the appendix).
file_summary <- tibble(
  item = c(
    "compressed_data_file",
    "ddi_codebook",
    "compressed_size_mib",
    "raw_extract_used",
    "processed_outputs_current",
    "rebuild_from_raw"
  ),
  value = c(
    if (has_raw_extract) basename(dat_gz_file) else NA_character_,
    if (has_raw_extract) basename(ddi_file) else NA_character_,
    if (has_raw_extract) round(file.info(dat_gz_file)$size / 1024^2, 1) else NA_real_,
    has_raw_extract,
    processed_current,
    rebuild_processed
  )
)
```

```{r build-or-load-aggregates}
if (rebuild_processed) {
  # Parse only the variables used for filtering, cohort assignment, weights,
  # and fertility counts.
  ipums_data <- read_ipums_micro(
    ddi,
    vars = analysis_vars,
    data_file = dat_gz_file,
    verbose = FALSE
  ) %>%
    mutate(across(all_of(analysis_vars), as.numeric))

  # --- Main series: age-first cohort assignment (membership defined by AGE within
  # a census). The 1950 cohort is a proxy from women aged 60-69 in 1960; direct
  # 1950 (women 50-59 in 1950) is held out for validation. Birth-year spans are
  # derived afterwards for labeling only.
  cohort_records <- ipums_data %>%
    filter(SEX == 2) %>%
    mutate(
      source_year = YEAR,
      cohort_source = case_when(
        YEAR == 1950 & between(AGE, 50, 59) ~ "direct_1950_validation",
        YEAR == 1960 & between(AGE, 60, 69) ~ "proxy_1950",
        YEAR == 1960 & between(AGE, 50, 59) ~ "main_1960",
        YEAR == 1970 & between(AGE, 50, 59) ~ "main_1970",
        YEAR == 1980 & between(AGE, 50, 59) ~ "main_1980",
        YEAR == 1990 & between(AGE, 50, 59) ~ "main_1990",
        TRUE ~ NA_character_
      ),
      Year = case_when(
        cohort_source %in% c("proxy_1950", "direct_1950_validation") ~ 1950,
        cohort_source == "main_1960" ~ 1960,
        cohort_source == "main_1970" ~ 1970,
        cohort_source == "main_1980" ~ 1980,
        cohort_source == "main_1990" ~ 1990,
        TRUE ~ NA_real_
      ),
      series = case_when(
        cohort_source == "direct_1950_validation" ~ "validation",
        !is.na(cohort_source) ~ "main",
        TRUE ~ NA_character_
      ),
      valid_chborn = CHBORN > 0 & CHBORN < 97,
      n_child = if_else(valid_chborn, pmin(CHBORN - 1, 12), NA_real_)
    ) %>%
    filter(!is.na(cohort_source))

  cohort_composition <- cohort_records %>%
    group_by(series, cohort_source, source_year, Year) %>%
    summarise(
      n_records = n(),
      age_min = min(AGE, na.rm = TRUE),
      age_max = max(AGE, na.rm = TRUE),
      birthyr_min = min(BIRTHYR[BIRTHYR > 1800], na.rm = TRUE),
      birthyr_max = max(BIRTHYR[BIRTHYR > 1800], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(cohort = paste0(birthyr_min, "-", birthyr_max)) %>%
    arrange(Year, source_year, cohort_source)

  cohort_records <- cohort_records %>%
    left_join(select(cohort_composition, cohort_source, cohort), by = "cohort_source")

  age_cohort_validation <- cohort_composition
  completed_fertility <- cohort_records
  valid_completed_fertility <- completed_fertility %>% filter(valid_chborn)
  main_fertility <- valid_completed_fertility %>% filter(series == "main")
  validation_1950 <- valid_completed_fertility %>%
    filter(cohort_source %in% c("proxy_1950", "direct_1950_validation"))

  # How much data the CHBORN validity rule excludes, by cohort source.
  chborn_universe <- completed_fertility %>%
    count(cohort_source, source_year, Year, cohort, valid_chborn, name = "n_records") %>%
    pivot_wider(
      names_from = valid_chborn,
      values_from = n_records,
      names_prefix = "valid_chborn_",
      values_fill = 0
    ) %>%
    mutate(
      total_records = valid_chborn_FALSE + valid_chborn_TRUE,
      valid_share = valid_chborn_TRUE / total_records
    ) %>%
    arrange(Year, source_year, cohort_source)

  cohort_means <- valid_completed_fertility %>%
    group_by(cohort_source, series, source_year, Year, cohort) %>%
    summarise(
      n_women = n(),
      n_women_weighted = sum(PERWT, na.rm = TRUE),
      mean_children = mean(n_child),
      mean_children_weighted = weighted.mean(n_child, PERWT, na.rm = TRUE),
      share_12plus = mean(n_child == 12),
      .groups = "drop"
    ) %>%
    arrange(series, Year, source_year, cohort_source)

  fertility_counts <- main_fertility %>%
    count(Year, cohort, n_child, name = "n_women") %>%
    arrange(Year, n_child)

  fertility_counts_weighted <- main_fertility %>%
    group_by(Year, cohort, n_child) %>%
    summarise(n_women_weighted = sum(PERWT, na.rm = TRUE), .groups = "drop") %>%
    arrange(Year, n_child)

  fertility_pmf <- fertility_counts %>%
    group_by(Year, cohort) %>%
    mutate(probability = n_women / sum(n_women)) %>%
    ungroup() %>%
    transmute(Year, cohort, children = n_child, probability)

  fertility_pmf_weighted <- fertility_counts_weighted %>%
    group_by(Year, cohort) %>%
    mutate(probability_weighted = n_women_weighted / sum(n_women_weighted)) %>%
    ungroup() %>%
    transmute(Year, cohort, children = n_child, probability_weighted)

  cohort_sample_sizes <- main_fertility %>%
    group_by(Year, cohort) %>%
    summarise(
      n_women = n(),
      n_women_weighted = sum(PERWT, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(Year)

  validation_1950_summary <- validation_1950 %>%
    group_by(cohort_source) %>%
    summarise(
      source_year = first(source_year),
      output_year = first(Year),
      n_women = n(),
      n_women_weighted = sum(PERWT, na.rm = TRUE),
      mean_children = mean(n_child),
      mean_children_weighted = weighted.mean(n_child, PERWT, na.rm = TRUE),
      share_12plus = mean(n_child == 12),
      .groups = "drop"
    ) %>%
    arrange(cohort_source)

  validation_1950_pmf <- validation_1950 %>%
    count(cohort_source, n_child, name = "n_women") %>%
    left_join(
      validation_1950 %>%
        group_by(cohort_source, n_child) %>%
        summarise(n_women_weighted = sum(PERWT, na.rm = TRUE), .groups = "drop"),
      by = c("cohort_source", "n_child")
    ) %>%
    group_by(cohort_source) %>%
    mutate(
      probability = n_women / sum(n_women),
      probability_weighted = n_women_weighted / sum(n_women_weighted)
    ) %>%
    ungroup() %>%
    arrange(cohort_source, n_child)

  # --- Common-universe analysis (separate frame; the main series is untouched).
  # Follow decade birth cohorts across the 1960-1990 censuses and, where the whole
  # cohort has reached age 50+, contrast the children-ever-born distribution under
  # two universes: all women vs ever-married. This isolates the effect of the
  # universe shift (1950/1960 asked only ever-married women; 1970-1990 all women).
  # MARST: 1-5 = ever-married, 6 = never-married, 9 = blank/missing (kept in the
  # all-women universe but excluded from the ever-married subset).
  band_records <- ipums_data %>%
    filter(SEX == 2, between(YEAR, 1960, 1990), between(BIRTHYR, 1890, 1939)) %>%
    mutate(
      band_start = 10L * (BIRTHYR %/% 10L),
      birth_band = paste0(band_start, "-", band_start + 9L),
      universe_year = YEAR,
      # Keep only cells where the whole decade cohort has reached age 50+.
      cohort_complete = (universe_year - (band_start + 9L)) >= 50L,
      valid_chborn = CHBORN > 0 & CHBORN < 97,
      n_child = if_else(valid_chborn, pmin(CHBORN - 1, 12), NA_real_),
      marital = case_when(
        MARST %in% 1:5 ~ "ever_married",
        MARST == 6 ~ "never_married",
        TRUE ~ "unknown"
      ),
      gq_household = GQ %in% c(1, 2, 5)
    ) %>%
    filter(cohort_complete, valid_chborn)

  band_long <- bind_rows(
    band_records %>% mutate(universe = "all_women"),
    band_records %>% filter(marital == "ever_married") %>% mutate(universe = "ever_married")
  )

  band_summary <- band_long %>%
    group_by(birth_band, universe_year, universe) %>%
    summarise(
      age_min = min(AGE),
      age_max = max(AGE),
      n_women = n(),
      n_women_weighted = sum(PERWT),
      mean_children = mean(n_child),
      mean_children_weighted = weighted.mean(n_child, PERWT),
      var_children = var(n_child),
      share_0 = mean(n_child == 0),
      share_12plus = mean(n_child == 12),
      gq_household_share = mean(gq_household),
      .groups = "drop"
    ) %>%
    arrange(birth_band, universe_year, universe)

  band_pmf <- band_long %>%
    count(birth_band, universe_year, universe, n_child, name = "n_women") %>%
    left_join(
      band_long %>%
        group_by(birth_band, universe_year, universe, n_child) %>%
        summarise(n_women_weighted = sum(PERWT), .groups = "drop"),
      by = c("birth_band", "universe_year", "universe", "n_child")
    ) %>%
    group_by(birth_band, universe_year, universe) %>%
    mutate(
      probability = n_women / sum(n_women),
      probability_weighted = n_women_weighted / sum(n_women_weighted)
    ) %>%
    ungroup() %>%
    arrange(birth_band, universe_year, universe, n_child)

  # Within-census contrast. The two universes coincide by construction before 1970
  # (never-married women were not in the CHBORN universe), so the contrast is
  # meaningful only for universe_year >= 1970.
  universe_contrast <- band_summary %>%
    select(birth_band, universe_year, universe, n_women, mean_children, var_children, share_0) %>%
    pivot_wider(
      names_from = universe,
      values_from = c(n_women, mean_children, var_children, share_0)
    ) %>%
    mutate(
      mean_shift = mean_children_all_women - mean_children_ever_married,
      share0_shift = share_0_all_women - share_0_ever_married,
      pct_never_married = 1 - n_women_ever_married / n_women_all_women
    )

  tvd_table <- band_pmf %>%
    select(birth_band, universe_year, universe, n_child, probability, probability_weighted) %>%
    pivot_wider(
      names_from = universe,
      values_from = c(probability, probability_weighted),
      values_fill = 0
    ) %>%
    group_by(birth_band, universe_year) %>%
    summarise(
      tvd = 0.5 * sum(abs(probability_all_women - probability_ever_married)),
      tvd_weighted = 0.5 * sum(abs(probability_weighted_all_women - probability_weighted_ever_married)),
      .groups = "drop"
    )

  universe_contrast <- universe_contrast %>%
    left_join(tvd_table, by = c("birth_band", "universe_year")) %>%
    arrange(birth_band, universe_year)

  # Early-cohort decomposition: the two cohorts seen ever-married at completed
  # fertility (1960) and all-women later (1970). Split the gap into the universe
  # effect (same census, adding never-married women) and the mortality/decade
  # effect (same ever-married universe, older age).
  pick_mean <- function(b, y, u) {
    v <- band_summary$mean_children[
      band_summary$birth_band == b &
        band_summary$universe_year == y &
        band_summary$universe == u
    ]
    if (length(v) == 1) v else NA_real_
  }
  early_cohort_followup <- tibble(
    birth_band = c("1890-1899", "1900-1909"),
    completed_year = c(1960, 1960),
    later_year = c(1970, 1970)
  ) %>%
    mutate(
      completed_em_mean = mapply(function(b, y) pick_mean(b, y, "ever_married"), birth_band, completed_year),
      later_em_mean = mapply(function(b, y) pick_mean(b, y, "ever_married"), birth_band, later_year),
      later_all_mean = mapply(function(b, y) pick_mean(b, y, "all_women"), birth_band, later_year),
      universe_effect = later_all_mean - later_em_mean,
      mortality_effect = later_em_mean - completed_em_mean
    )

  # Self-check: the 1900-1909 / 1960 ever-married mean should be close to the main
  # 1960 series mean, and 1890-1899 / 1960 close to the proxy-1950 mean.
  message(
    "Self-check (common-universe vs main series): ",
    "1900-1909/1960 ever-married mean = ", round(pick_mean("1900-1909", 1960, "ever_married"), 3),
    " (main_1960 ~ 2.429); ",
    "1890-1899/1960 ever-married mean = ", round(pick_mean("1890-1899", 1960, "ever_married"), 3),
    " (proxy_1950 ~ 2.818)."
  )

  # Write the aggregate CSVs.
  write_csv(cohort_sample_sizes, file.path(processed_dir, "ipums_cohort_sample_sizes.csv"))
  write_csv(fertility_counts, file.path(processed_dir, "ipums_fertility_counts_by_cohort.csv"))
  write_csv(fertility_pmf, file.path(processed_dir, "ipums_fertility_pmf_by_cohort.csv"))
  write_csv(fertility_counts_weighted, file.path(processed_dir, "ipums_fertility_counts_by_cohort_weighted.csv"))
  write_csv(fertility_pmf_weighted, file.path(processed_dir, "ipums_fertility_pmf_by_cohort_weighted.csv"))
  write_csv(validation_1950_summary, file.path(processed_dir, "validation_1950_direct_vs_proxy.csv"))
  write_csv(validation_1950_pmf, file.path(processed_dir, "validation_1950_direct_vs_proxy_pmf.csv"))
  write_csv(age_cohort_validation, file.path(processed_dir, "ipums_age_cohort_validation.csv"))
  write_csv(chborn_universe, file.path(processed_dir, "ipums_chborn_universe_by_cohort.csv"))
  write_csv(cohort_means, file.path(processed_dir, "ipums_cohort_fertility_checks.csv"))
  write_csv(band_summary, file.path(processed_dir, "ipums_cohort_universe_summary.csv"))
  write_csv(band_pmf, file.path(processed_dir, "ipums_cohort_universe_pmf.csv"))
  write_csv(universe_contrast, file.path(processed_dir, "ipums_universe_contrast_by_cohort.csv"))
  write_csv(early_cohort_followup, file.path(processed_dir, "ipums_early_cohort_followup.csv"))
  write_csv(
    tibble(
      processing_version = processing_version,
      cohort_definition = "main series: age-defined cohorts (50-59; proxy-1950 from 60-69 in 1960; direct-1950 validation-only); CHBORN harmonized to 12+. Common-universe analysis: decade birth cohorts followed across 1960-1990 at completed ages, all-women vs ever-married (MARST 1-5).",
      rebuilt_at = as.character(Sys.time())
    ),
    file.path(processed_dir, "ipums_preprocess_metadata.csv")
  )

  fertility_pmf %>%
    group_split(Year) %>%
    lapply(function(dat) {
      year <- unique(dat$Year)
      write_csv(
        dat %>% transmute(children, probability),
        file.path(processed_dir, paste0("fertility_pmf_", year, ".csv"))
      )
    })

  rm(
    ipums_data, cohort_records, cohort_composition, completed_fertility,
    valid_completed_fertility, main_fertility, validation_1950,
    band_records, band_long, tvd_table
  )
} else {
  # Load the committed aggregates without reparsing restricted person-level data.
  rd <- function(f) read_csv(file.path(processed_dir, f), show_col_types = FALSE)
  cohort_sample_sizes <- rd("ipums_cohort_sample_sizes.csv")
  fertility_counts <- rd("ipums_fertility_counts_by_cohort.csv")
  fertility_pmf <- rd("ipums_fertility_pmf_by_cohort.csv")
  fertility_counts_weighted <- rd("ipums_fertility_counts_by_cohort_weighted.csv")
  fertility_pmf_weighted <- rd("ipums_fertility_pmf_by_cohort_weighted.csv")
  validation_1950_summary <- rd("validation_1950_direct_vs_proxy.csv")
  validation_1950_pmf <- rd("validation_1950_direct_vs_proxy_pmf.csv")
  age_cohort_validation <- rd("ipums_age_cohort_validation.csv")
  chborn_universe <- rd("ipums_chborn_universe_by_cohort.csv")
  cohort_means <- rd("ipums_cohort_fertility_checks.csv")
  band_summary <- rd("ipums_cohort_universe_summary.csv")
  band_pmf <- rd("ipums_cohort_universe_pmf.csv")
  universe_contrast <- rd("ipums_universe_contrast_by_cohort.csv")
  early_cohort_followup <- rd("ipums_early_cohort_followup.csv")
  preprocess_metadata <- rd("ipums_preprocess_metadata.csv")
  if (!identical(preprocess_metadata$processing_version[[1]], processing_version)) {
    stop(
      "Processed files were built with processing version ",
      preprocess_metadata$processing_version[[1]],
      "; expected ", processing_version,
      ". Re-render with KINCOUNTS_REBUILD_IPUMS=true.",
      call. = FALSE
    )
  }
}

# Confirm the complete aggregate output set is available after either branch.
written_files <- tibble(
  file = processed_files,
  status = if_else(file.exists(processed_paths), "available", "missing")
)
if (any(written_files$status == "missing")) {
  stop(
    "Missing processed files: ",
    paste(written_files$file[written_files$status == "missing"], collapse = ", "),
    call. = FALSE
  )
}
```

## The children-ever-born question and its universe

`CHBORN` records the number of **children ever born alive** to a woman. It counts live
births only — **stillbirths are excluded**, as are adopted children and stepchildren.

For a multi-decade comparison, the central complication is that *who was asked the
question* — the variable's **universe** — changed across censuses:

| Census | Universe (who was asked) |
|---|---|
| 1950 | ever-married women, **and only the 1-in-6 sample line** |
| 1960 | ever-married women |
| 1970 | all women age 15+ |
| 1980 | all women age 15+ |
| 1990 | all women age 15+ |

A woman not in the universe was not asked the question and has no recorded value (IPUMS
code `00`), which is distinct from a recorded count of zero. Such records are excluded
from all fertility distributions. Two consequences follow.

**Consequence 1 — coverage (1950).** Because 1950 asked only ever-married *sample-line*
women, most 1950 women have no `CHBORN` value. The figure below shows, for each
census/age group, the share of women who were asked the question: direct-1950 is near
30% (only ever-married sample-line women); the 1960 census — the source of both the 1960
cohort and the 1950 proxy — is near 92% (ever-married women only); and the 1970–1990
censuses are at 100% (all women). This is why the 1950 cohort is taken from a **proxy**
(the same birth cohort observed in 1960) rather than directly (see *Defining the analytic
sample*).

```{r fig-coverage}
#| fig-cap: "Of the women in each census/age group, the share who were asked the children-ever-born question (i.e., were in the variable's universe). Colour indicates whom each census asked: in 1950 only ever-married women on the 1-in-6 sample line (~30%); in 1960 only ever-married women (~92%); from 1970 all women (~100%). The low 1950 coverage is why the 1950 cohort is taken from the 1960 proxy."
#| fig-width: 8.5
#| fig-height: 3.8
coverage_labels <- tibble(
  cohort_source = c("direct_1950_validation", "proxy_1950", "main_1960",
                    "main_1970", "main_1980", "main_1990"),
  slice = c("1950 cohort (direct)", "1950 cohort (proxy)", "1960 cohort",
            "1970 cohort", "1980 cohort", "1990 cohort"),
  who_asked = c("Ever-married, 1-in-6 sample line", "Ever-married women",
                "Ever-married women", "All women", "All women", "All women")
)

chborn_universe %>%
  left_join(coverage_labels, by = "cohort_source") %>%
  mutate(label = paste0(slice, ", ", source_year, " census")) %>%
  ggplot(aes(x = reorder(label, valid_share), y = valid_share, fill = who_asked)) +
  geom_col() +
  geom_text(aes(label = scales::label_percent(accuracy = 1)(valid_share)),
            hjust = -0.15, size = 3.4) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = scales::label_percent(), limits = c(0, 1.1),
                     expand = expansion(mult = c(0, 0))) +
  scale_fill_manual(values = coverage_palette) +
  labs(x = NULL, y = "Share of women in the age group who were asked CHBORN",
       fill = "Who the census asked",
       title = "Who was asked the children-ever-born question")
```

**Consequence 2 — the universe shift (comparability).** The early cohorts describe
ever-married women only; the later cohorts describe all women. A raw cross-cohort series
therefore mixes two populations: the 1970–1990 samples include never-married (largely
childless) women that the 1950 and 1960 samples cannot. The only universe common to every
census is ever-married women, which can be reconstructed in any year from marital status
(`MARST`); this motivates the common-universe analysis below.

## Defining the analytic sample

The analytic population is women (`SEX == 2`) observed at completed fertility. Cohort
membership for the **main series** is defined **by age within a census** — women aged
50–59 — not by birth year, so nobody in the age window is dropped because their birthday
had not yet occurred by the April-1 census date. Birth-year spans are *derived* from the
data afterwards, for labeling only.

| Output year | Source census | Age window | Derived birth years* | Role |
|---:|---:|---:|---|---|
| 1950 | 1960 | 60–69 | ~1890–1900 | Main proxy for the 1950 cohort |
| 1960 | 1960 | 50–59 | ~1900–1910 | Main series |
| 1970 | 1970 | 50–59 | ~1910–1920 | Main series |
| 1980 | 1980 | 50–59 | ~1920–1930 | Main series |
| 1990 | 1990 | 50–59 | ~1930–1940 | Main series |
| 1950 | 1950 | 50–59 | ~1890–1900 | Direct-1950 validation only |

\*A 10-year age window spans about 11 birth years (the April-1 birthday effect), so
adjacent cohorts can share a boundary birth year. The exact span is derived from the
data and shown below.

The **1950 cohort uses the 1960 proxy** (the 1890–1900 birth cohort observed at ages
60–69 in 1960), because — per *Consequence 1* above — direct 1950 retains only ~30% of
women. Direct-1950 is computed and reported only as a validation check.

```{r tbl-composition}
age_cohort_validation %>%
  select(series, cohort_source, source_year, Year, cohort, age_min, age_max, n_records) %>%
  kable(caption = "Cohort composition: age range and derived birth-year span for each age-defined cohort slice (no records are dropped by age).")
```

## CHBORN coding and harmonization

`ipumsr` returns the raw IPUMS codes. Code `00` is *not applicable* (not in universe —
**not** zero children); zero children is code `01`, one child `02`, and so on. Per the
IPUMS user note, children ever born is the code minus one, after dropping `00`. We keep
codes `01`–`61`, set `n_child = CHBORN − 1`, and **cap at 12** (`12+`) so all cohorts
share one top category: the 1960–1990 samples top-code at `12+` (code `13`), while 1950
resolves higher (to `25+`, code `26`).

```{r tbl-coding}
value_labels %>%
  filter(var_name %in% c("SEX", "CHBORN", "MARST")) %>%
  kable(caption = "DDI value labels for the variables that drive the coding rules.")

tibble(
  raw_code = c("0", "1", "2-12", "13", "14-61", "97, 98, 99"),
  ddi_meaning = c(
    "N/A (not in universe)",
    "No children",
    "1 to 11 children",
    "12 children; 12+ in 1960-1990",
    "13 to 60 children (1950 coding)",
    "Unknown, illegible, or missing"
  ),
  harmonized_use = c(
    "Drop (not zero children)",
    "n_child = 0",
    "n_child = CHBORN - 1",
    "n_child = 12 (the 12+ bucket)",
    "n_child = 12 (the 12+ bucket)",
    "Drop"
  )
) %>%
  kable(caption = "Harmonized CHBORN coding used in all processed outputs.")
```

## Putting cohorts on a common universe

We treat fertility as effectively complete by age 50; the 50–59 age window is chosen to
capture completed fertility while limiting the mortality selection that affects older
ages. Under this assumption, differences in a fixed birth cohort's measured
children-ever-born distribution across successive censuses reflect measurement —
mortality selection, migration, response error, and universe or top-code differences —
rather than a change in fertility. We therefore estimate the universe effect by comparing
the two universes *within a single census*, and use the cross-census comparison only to
gauge the magnitude of these measurement differences.

We follow each decade birth cohort across the 1960–1990 censuses, restricting to cells
where the whole cohort has reached age 50+, and within each census compute the
children-ever-born distribution two ways: **all women** vs **ever-married** (`MARST 1–5`).
Before 1970 the two coincide by construction (never-married women were not in the
universe), so the contrast is informative only for 1970 onward.

```{r fig-trajectory}
#| fig-cap: "Mean children ever born for each decade birth cohort, followed across censuses, under the two universes. The series coincide in 1960 (only ever-married women were asked) and separate from 1970 onward (all women asked); the gap between the series is the universe effect, while the drift along each line reflects mortality selection and decade differences from observing a cohort at older ages."
#| fig-width: 9
#| fig-height: 6
band_summary %>%
  mutate(Universe = recode(universe, all_women = "All women", ever_married = "Ever-married")) %>%
  ggplot(aes(universe_year, mean_children, color = Universe, group = Universe)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~ birth_band) +
  scale_color_manual(values = universe_palette) +
  scale_x_continuous(breaks = c(1960, 1970, 1980, 1990)) +
  labs(x = "Census year", y = "Mean children ever born",
       title = "Same birth cohorts under all-women vs ever-married universes")
```

```{r fig-early-pmf}
#| fig-cap: "Children-ever-born PMFs for the two early cohorts in the first census where all women were asked (1970). Ever-married (the only universe available for these cohorts at completed fertility in 1960) vs all women. Including never-married women adds mass at zero and lowers the mean."
#| fig-width: 9
#| fig-height: 4.5
band_pmf %>%
  filter(birth_band %in% c("1890-1899", "1900-1909"), universe_year == 1970) %>%
  mutate(Universe = recode(universe, all_women = "All women", ever_married = "Ever-married")) %>%
  ggplot(aes(n_child, probability, fill = Universe)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  facet_wrap(~ birth_band) +
  scale_fill_manual(values = universe_palette) +
  scale_x_continuous(breaks = 0:12) +
  labs(x = "Children ever born (12 = 12+)", y = "Probability",
       title = "All-women vs ever-married PMFs, early cohorts, 1970")
```

```{r fig-shift}
#| fig-cap: "The universe effect: how much the mean falls when the universe is widened from ever-married to all women, by cohort and census (1970+ only, where the two universes differ)."
#| fig-width: 8
#| fig-height: 4.5
universe_contrast %>%
  filter(universe_year >= 1970) %>%
  ggplot(aes(factor(universe_year), mean_shift, fill = birth_band)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  scale_fill_manual(
    values = setNames(
      viridis(length(unique(universe_contrast$birth_band)), option = "D", begin = 0.12, end = 0.88),
      sort(unique(universe_contrast$birth_band))
    )
  ) +
  labs(x = "Census year", y = "Mean shift (all women − ever-married)",
       fill = "Birth cohort",
       title = "Effect of widening the universe to all women")
```

The contrast and decomposition tables accompany the figures:

```{r tbl-universe}
universe_contrast %>%
  filter(universe_year >= 1970) %>%
  transmute(
    birth_band, universe_year,
    mean_all = mean_children_all_women,
    mean_em = mean_children_ever_married,
    mean_shift, share0_shift, pct_never_married, tvd
  ) %>%
  kable(digits = 3, caption = "All-women vs ever-married contrast by cohort and census (1970+). mean_shift = all − ever-married; pct_never_married = never-married share of the all-women cell; tvd = total-variation distance between the two PMFs.")

early_cohort_followup %>%
  kable(digits = 3, caption = "Early-cohort decomposition. universe_effect (all − ever-married in the later census) isolates the effect of adding never-married women; mortality_effect (ever-married, later − completion census) isolates survivor selection and decade artifacts.")
```

**Interpretation.** The within-census all-women vs ever-married contrast (`mean_shift`,
`tvd`, and the 1970 PMFs) is the preferred estimate: it compares the same survivors in the
same census, with appropriate `PERWT` weighting. Differences across censuses — the drift
along each trajectory — bound the combined measurement difference rather than a change in
fertility. Key caveats:

- **Mortality selection.** Never-married women have higher old-age mortality, so the
  all-women pool drifts *toward* ever-married at older ages — the universe gap is
  *attenuated* at older ages. The least-contaminated estimate is the contrast at the
  **youngest census** in which a cohort is observed under the all-women universe.
- **Never-married fertility is a lower bound.** Historically nonmarital births are
  under-reported, so the all-women mean understates true all-women fertility and the gap
  is, if anything, overstated.
- **Weighting.** `PERWT` is appropriate for 1970–1990 (CHBORN is universal there), so the
  within-census contrast is unaffected by the weighting issue below. 1950 is excluded from
  this analysis (its sample-line weight `SLWT` is absent), and the 1960 cells are
  ever-married by construction.
- **Migration / nativity.** Birth cohorts are not closed across censuses; `NATIVITY`/`BPL`
  are absent, so cross-census comparisons cannot be restricted to the native-born — a
  further reason to rely on the within-census contrast, which compares a single
  population. `gq_household_share` is reported per cell as a coarse check on
  institutionalization at older ages.

In sum, the pre-1970 cohorts (1950 and 1960) are observed only for ever-married women,
which raises their measured completed fertility by roughly 0.1–0.2 children relative to
an all-women basis. This gap is small next to the Depression-era decline in completed
fertility across the 1901–1920 birth cohorts, and it is reported as a sensitivity rather
than applied to the main series.

## Aggregate fertility distributions

The main-series outputs — completed-fertility distributions for women 50–59, with the
1950 proxy — are the inputs read by the downstream model and simulator. Means are
reported unweighted, the form those downstream steps require; `PERWT`-weighted versions
are written alongside as a sensitivity.

```{r fig-main-means}
#| fig-cap: "Mean children ever born by main-series cohort (women 50-59; 1950 is the 1960 proxy). The dip for the 1901-1920 cohorts and the rise for the baby-boom mothers are demographically expected."
#| fig-width: 7
#| fig-height: 4
cohort_means %>%
  filter(series == "main") %>%
  ggplot(aes(Year, mean_children)) +
  geom_line(linewidth = 0.8, color = main_series_color) +
  geom_point(size = 2.5, color = main_series_color) +
  geom_text(aes(label = round(mean_children, 2)), vjust = -0.9, size = 3.3) +
  scale_x_continuous(breaks = c(1950, 1960, 1970, 1980, 1990)) +
  labs(x = "Cohort (census year)", y = "Mean children ever born",
       title = "Completed fertility by cohort")
```

```{r fig-main-pmf}
#| fig-cap: "Fertility distributions (children ever born, 12 = 12+) for each main-series cohort."
#| fig-width: 9
#| fig-height: 6
fertility_pmf %>%
  ggplot(aes(children, probability)) +
  geom_col(fill = main_series_color, width = 0.8) +
  facet_wrap(~ Year) +
  scale_x_continuous(breaks = 0:12) +
  labs(x = "Children ever born (12 = 12+)", y = "Probability",
       title = "Fertility PMFs by cohort")
```

```{r tbl-main}
cohort_sample_sizes %>%
  kable(digits = 0, caption = "Main-series cohort sample sizes (unweighted and PERWT-weighted).")

cohort_means %>%
  filter(series == "main") %>%
  transmute(Year, cohort, n_women, mean_children, mean_children_weighted, share_12plus) %>%
  kable(digits = 3, caption = "Main-series mean children ever born and 12+ shares (unweighted and weighted).")
```

## Validation: direct vs proxy 1950

The 1950 cohort is represented by the 1960 proxy. As a check, we compare it against the
held-out direct-1950 records (women 50–59 in the 1950 census). The unweighted
distributions are nearly identical, which corroborates the proxy. (The weighted
direct-1950 number is unreliable — 1950 `CHBORN` is a sample-line variable and the
correct weight `SLWT` is not in the extract — so we rely on the unweighted comparison,
and tail differences above 12 are affected by the top-code asymmetry noted in the coding
section.)

```{r fig-validation}
#| fig-cap: "Children-ever-born PMFs (unweighted): direct 1950 (women 50-59 in 1950) vs the 1960 proxy (women 60-69 in 1960). Near-identical shapes corroborate using the proxy."
#| fig-width: 8
#| fig-height: 4.5
validation_1950_pmf %>%
  mutate(Source = recode(cohort_source,
                         direct_1950_validation = "Direct 1950",
                         proxy_1950 = "Proxy (1960)")) %>%
  ggplot(aes(n_child, probability, fill = Source)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  scale_fill_manual(values = validation_palette) +
  scale_x_continuous(breaks = 0:12) +
  labs(x = "Children ever born (12 = 12+)", y = "Probability",
       title = "1950 cohort: direct vs proxy")
```

```{r tbl-validation}
validation_1950_summary %>%
  kable(digits = 3, caption = "Direct 1950 vs proxy 1950 summary. Unweighted means agree closely; the weighted direct-1950 number is approximate (no SLWT).")
```

## Limitations

- **12+ top-code.** All outputs harmonize the child-count support to `0, 1, …, 11, 12+`.
  1960–1990 already top-code at `12+`, but 1950 resolves to `25+`, so the cap discards
  tail detail for the (validation-only) 1950 series and for older cohorts seen in 1950.
- **Weighting.** Main outputs are unweighted; `PERWT`-weighted versions are provided as a
  sensitivity. The 1950 sample-line weight `SLWT` is **absent**, so any 1950 weighted
  figure is approximate; 1950 is therefore kept out of the weighted cross-cohort series.
- **Extract gaps.** A future extract should add **`SLWT`** (correct 1950 weighting),
  **`NATIVITY`/`BPL`** (restrict to native-born / measure cohort compositional drift
  across censuses), and **`QCHBORN`** (filter allocated/imputed values). Each would close
  a caveat that the current data cannot resolve.
- **Sharing.** The restricted IPUMS records, person-level weights, and raw fixed-width
  data remain local-only and are not committed; only the aggregate CSVs are.

## Appendix: reproducibility and outputs

This page is processed-data-first: a normal render loads the committed aggregate CSVs
from `data/processed/` and does not look for or open restricted raw IPUMS files. Set
`KINCOUNTS_REBUILD_IPUMS=true` to explicitly rebuild the aggregates from a local raw
extract.

```{r tbl-appendix}
kable(file_summary, caption = "Local restricted IPUMS files and cache state for this render.")
kable(ddi_variables, caption = "Parsed variables and their fixed-width positions in the DDI.")
kable(
  tibble(rebuild_from_raw = rebuild_processed, processed_file_count = length(processed_files)),
  caption = "Preprocessing cache status for this render."
)
kable(written_files, caption = "Aggregate processed files available in data/processed.")
kable(output_inventory, caption = "Full inventory of aggregate output files.")
```
