---
title: "Simulate the number of relatives under different fertility and sibling models"
author: "Junhui He"
date: today
---

# Simulate the number of relatives

We simulate the number of relatives for each individual using fertility and sibling distributions.

```{r setup, message = FALSE, warning = FALSE, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE, fig.width = 9, fig.height = 7)

library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(MASS)     ## glm.nb
library(pscl)     ## zeroinfl
library(knitr)    ## kable

output_path <- if (dir.exists("output")) "output" else file.path("..", "output")
```

## Simulate the relative counts for the focal individual

```{r simulate-relatives}
# The fertility distribution modeled as a zero-inflated negative binomial distribution.
simulate_relatives_zinb <- function(mus, sizes, pis,
                                    n_sim = 10000,
                                    max_kids = 12) {
  draw_fert <- function(n, mu, size, pi) {
    ifelse(runif(n) < pi, 0L,
           as.integer(pmin(rnbinom(n, size, mu = mu), max_kids)))
  }
  
  draw_sibs <- function(n, mu, size) {
    as.integer(pmin(rnbinom(n, size + 1, mu = mu * (size + 1) / size), max_kids - 1))
  }
  # 1. Parents and grandparents are fixed at 2 and 4, respectively.
  parents <- rep(2L, n_sim)
  grandparents <- rep(4L, n_sim)

  # 2. Siblings for the focal individual
  siblings <- draw_sibs(n = n_sim, mu = mus[2], size = sizes[2])
  
  # 3. Children of the focal individual
  children <- draw_fert(n = n_sim, mu = mus[1], size = sizes[1], pi = pis[1])

  # 4. Aunts & uncles (no subtraction needed)
  aunts_uncles <- purrr::map_int(parents, function(n_par) {
    as.integer(sum(draw_sibs(n_par, mu = mus[3], size = sizes[3])))
  })

  # 5. Nieces & nephews: for each sibling, simulate their fertility and sum
  nieces_nephews <- purrr::map_int(siblings, function(n_sib) {
    if (n_sib == 0L) 0L else as.integer(sum(draw_fert(n_sib, mu = mus[1], size = sizes[1], pi = pis[1])))
  })

  # 6. Cousins: for each aunt/uncle, simulate their fertility and sum
  cousins <- purrr::map_int(aunts_uncles, function(n_au) {
    if (n_au == 0L) 0L else as.integer(sum(draw_fert(n_au, mu = mus[2], size = sizes[2], pi = pis[2])))
  })

  tibble::tibble(
    Parents    = parents,
    Grandparents = grandparents,
    Siblings    = siblings,
    Children    = children,
    AuntUncles = aunts_uncles,
    NieceNephews = nieces_nephews,
    Cousins     = cousins
  )
}

# The fertility distribution modeled as a Poisson distribution.
simulate_relatives_poisson <- function(mus, 
                                       n_sim = 10000,
                                       max_kids = 12) {
  draw_fert <- function(n, mu) {
    as.integer(pmin(rpois(n, lambda = mu), max_kids))
  }
  
  draw_sibs <- function(n, mu, size) {
    as.integer(pmin(rpois(n, lambda = mu), max_kids - 1))
  }
  # 1. Parents and grandparents are fixed at 2 and 4, respectively.
  parents <- rep(2L, n_sim)
  grandparents <- rep(4L, n_sim)

  # 2. Siblings for the focal individual
  siblings <- draw_sibs(n = n_sim, mu = mus[2])
  
  # 3. Children of the focal individual
  children <- draw_fert(n = n_sim, mu = mus[1])

  # 4. Aunts & uncles (no subtraction needed)
  aunts_uncles <- purrr::map_int(parents, function(n_par) {
    as.integer(sum(draw_sibs(n_par, mu = mus[3])))
  })

  # 5. Nieces & nephews: for each sibling, simulate their fertility and sum
  nieces_nephews <- purrr::map_int(siblings, function(n_sib) {
    if (n_sib == 0L) 0L else as.integer(sum(draw_fert(n_sib, mu = mus[1])))
  })

  # 6. Cousins: for each aunt/uncle, simulate their fertility and sum
  cousins <- purrr::map_int(aunts_uncles, function(n_au) {
    if (n_au == 0L) 0L else as.integer(sum(draw_fert(n_au, mu = mus[2])))
  })

  tibble::tibble(
    Parents    = parents,
    Grandparents = grandparents,
    Siblings    = siblings,
    Children    = children,
    AuntUncles = aunts_uncles,
    NieceNephews = nieces_nephews,
    Cousins     = cousins
  )
}

# The fertility distribution fixed at the mean size.
simulate_relatives_fixed <- function(ferts, sibs) {
  # 1. Parents and grandparents are fixed at 2 and 4, respectively.
  parents <- 2
  grandparents <- 4

  # 2. Siblings for the focal individual
  siblings <- sibs[2]
  
  # 3. Children of the focal individual
  children <- ferts[1]

  # 4. Aunts & uncles (no subtraction needed)
  aunts_uncles <- parents * sibs[3]

  # 5. Nieces & nephews: for each sibling, simulate their fertility and sum
  nieces_nephews <- siblings * ferts[1]

  # 6. Cousins: for each aunt/uncle, simulate their fertility and sum
  cousins <- aunts_uncles * ferts[2]

  tibble::tibble(
    Parents    = parents,
    Grandparents = grandparents,
    Siblings    = siblings,
    Children    = children,
    AuntUncles = aunts_uncles,
    NieceNephews = nieces_nephews,
    Cousins     = cousins
  )
}
```

## Fertility and sibling parameters based on the IPUMS data

```{r ipums-parameters}
# Read the fitted parameters and empirical summaries written by fert_model.qmd
# instead of hard-coding them, so this notebook stays in sync with the data.
fert_params <- read_csv(file.path(output_path, "fertility_parameters.csv"), show_col_types = FALSE)
fert_estimation <- read_csv(file.path(output_path, "fertility_estimation.csv"), show_col_types = FALSE)

# Generations used below, in order: focal individual, parent, grandparent.
gen_years <- c(1990, 1970, 1950)

zinb_params <- fert_params %>% filter(Model == "ZINB")
pois_params <- fert_params %>% filter(Model == "Poisson")
sib_estimation <- fert_estimation %>% filter(distribution == "sibling")

mus_zinb <- zinb_params$mu[match(gen_years, zinb_params$Year)]
sizes_zinb <- zinb_params$size[match(gen_years, zinb_params$Year)]
pis_zinb <- zinb_params$pi0[match(gen_years, zinb_params$Year)]
mus_poisson <- pois_params$mu[match(gen_years, pois_params$Year)]
mean_ferts <- mus_poisson
mean_sibs <- sib_estimation$mean_data[match(gen_years, sib_estimation$Year)]

n_sim <- 100000
```

## Simulate relative counts for each fertility and sibling model

```{r simulate-relative-counts}
relative_counts = list()
relative_counts[[1]] = simulate_relatives_fixed(
  ferts = mean_ferts,
  sibs = mean_sibs
)
relative_counts[[2]] = simulate_relatives_poisson(
  mus = mus_poisson,
  n_sim = n_sim
)
relative_counts[[3]] = simulate_relatives_zinb(
  mus = mus_zinb,
  sizes = sizes_zinb,
  pis = pis_zinb,
  n_sim = n_sim
)
```

## Summarize mean and variance of relative counts

```{r summarize-relative-counts}
relative_model_names <- c("fixed", "poisson", "zinb")

relative_summary <- Map(function(dat, model_name) {
  tibble(
    Relative = names(dat),
    mean = as.numeric(colMeans(dat)),
    variance = if (model_name == "fixed") {
      rep(0, ncol(dat))
    } else {
      as.numeric(sapply(dat, var))
    },
    model = model_name
  )
}, relative_counts, relative_model_names) %>%
  bind_rows()

relative_summary_table <- relative_summary %>%
  pivot_wider(
    names_from = model,
    values_from = c(mean, variance),
    names_glue = "{.value}_{model}"
  ) %>%
  dplyr::select(
    Relative,
    mean_fixed,
    mean_poisson,
    mean_zinb,
    variance_fixed,
    variance_poisson,
    variance_zinb
  )

write.csv(relative_summary_table, file.path(output_path, "relative_summary_table.csv"), row.names = FALSE)

kable(
  relative_summary_table,
  digits = 3,
  caption = "Mean and variance of relative counts by fertility and sibling model."
)
```

## Visualize the relative count distributions for each model

```{r visualize-relative-counts, fig.width = 10, fig.height = 4}
relative_distribution <- bind_rows(
  relative_counts[[2]] %>% mutate(Model = "Poisson"),
  relative_counts[[3]] %>% mutate(Model = "ZINB")
) %>%
  pivot_longer(
    cols = -Model,
    names_to = "Relative",
    values_to = "Count"
  ) %>%
  filter(!Relative %in% c("Parents", "Grandparents")) %>%
  mutate(
    Relative = recode(
      Relative,
      "Siblings" = "Siblings",
      "Children" = "Children",
      "AuntUncles" = "Aunts and uncles",
      "NieceNephews" = "Nieces and nephews",
      "Cousins" = "First cousins"
    ),
    Relative = factor(
      Relative,
      levels = c(
        "Siblings", "Children", "Aunts and uncles",
        "Nieces and nephews", "First cousins"
      )
    )
  )

ggplot(relative_distribution, aes(x = Model, y = Count, fill = Model)) +
  geom_boxplot(width = 0.65, alpha = 0.85, outlier.size = 0.35) +
  facet_wrap(~ Relative, scales = "free_y", nrow = 1) +
  scale_fill_manual(values = c("Poisson" = "#0072B2", "ZINB" = "#D55E00")) +
  labs(
    # title = "Relative Count Distributions Under Poisson and ZINB Models",
    x = NULL,
    y = "Relative count",
    fill = "Model"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

ggsave(file = file.path(output_path, "relative_count_distributions.pdf"), width = 10, height = 4)
```
